Software architectures to integrate workflow engines in science gateways

### 2016-05-14 paper.tex
Removed all the closed notes (#21)

\title{Software architectures to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures? what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'. They are just software architectures :) The point of the paper is to describe them properly. It's a general formulation to cover any kind of software architecture that is used to integrated workflow engines in science gateways. Does it make sense?}}

@@

\begin{abstract}
  \closednote{sil}{i find this a dry start. what about an intro sentence?
    Workflow engines are critical for the efficient and transparent
    exploitation of distributed infrastructures in the ecosystem of
    tools and services offered by science gateways.}{Tristan}{Done, thanks.}  Workflow engines
  are critical for the efficient and transparent exploitation of
  distributed infrastructures in the ecosystem of tools and services
  offered by science gateways.  We investigate six software
  architectures commonly used to integrate workflow engines into
  science gateways. In \emph{tight integration}, the workflow engine
  shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow import}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and support for meta-workflows and fine-grained
  debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}.\closednote{sil}{what are the insights for wf engine developers?. this is
    not clear in the discussion and conclusions. i think the focus
    lies more on the integration and not on one or the other}{Tristan}{Fixed by your next note.} Tight
  integration and task encapsulation are the easiest to integrate and
  the most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be
    made more concise and poignant}{Tristan}{Added a description of
    the architectures, following @pbellec's suggestion.}
  \closednote{sil}{what about closing with this sentence from above: These
      results provide insights for science gateway architects and
      workflow engine developers. }{Tristan}{Done, thanks.}These results provide insights for
		science gateway architects and developers.
\end{abstract}

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.} or  redundancy detection across
analyses to avoid re-computation. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), \closednote{Naj}{I think and is redundant here, you can replace is with a comma}{Tristan}{yes!} using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} 
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute
applications. \closednote{Naj}{provide examples?}{Tristan}{Examples
  are mentioned in the last paragraph of this section. Moved it next.}

@@

Robustness is highly desirable as well since it is key to a
good user experience. Simple architectures facilitate the
implementation effort towards robust interactions which, in turn, have
a positive impact on characteristics such as gateway predictability,
transparency, reliability, traceability and
reproducibility. \closednote{Marc-Etienne}{I find that something is
  weak in this last paragraph. Robustness will help with transparency,
  sure, but it seems only a small facet.  Also, the text seem to
  timidly favor architectures with less components/interactions
  without much explanation... Complex systems can be robust, but at
  greater costs...  Here is one: "Architectures showing reduced system
  complexity will greatly facilitate the implementation effort towards
  robust interactions which, in turn, have a positive impact on
  characteristics such as gateway predictability, transparence,
  general reliability and results traceability and reproducibility;
  characteristics that are also key to a good user
  experience."}{Tristan}{Agreed, fixed as
  suggested.}\closednote{Tristan}{Work on the this sentence again.}{Tristan}{done.}

Other specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, or social tools among users. \closednote{Naj}{I think using etc after such as, for example and e.g. is not correct. --sorry for being pedantic}{Tristan}{You are completely right, this is now fixed.}

@@

\closednote{Sil}{i changed above a few occurrences of "diagrammed". i know it is correct, but i am not used to seeing it. this is a matter of style, just forget my changes below then}{Tristan}{That's perfect, thanks!}

@@

\closednote{Marc-e}{A
  few things bother me a bit with the notation fig., making it harder
  to assimilate 1- The interactions are padded in the middle of other
  stuff (Administrator, Data) 2- The Software component example says
  "Science gateway", should it not be empty?  3- The red arrow to
  explain the use of red; it looks like a very specific type of
  interactions instead of the desired "any interactions or components
  can be red meaning...", especially with 'a' and 'd' which the reader
  does not know the meaning yet.  Maybe a little red arrow and square
  with "Red used when specific to workflow execution"?  I don't have a
  great solution...  Red has special meaning, maybe remove it from the
  other pics (workflow eng, tasks) as it creates lots of visual
  interference when looking for red?  4- Order in text jumps all over
  the fig notations. Explanations are given for, in order: Components,
  workflow engine, interactions, abstract interactions, red color,
  tasks, data (no Administrator?)...  Could re-ordering the fig fix
  this? A figure figure legend could go a long way too (are there some
  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}
\closednote{Naj}{I propose a table of definitions.}{Tristan}{Added
  information in Figure 1 and removed the corresponding text.}

@@

  with heterogeneous interfaces. \closedanswerednote{Marc-Etienne}{this -c2- can be also trigerred
    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?}
    {Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
    some operations across the infrastructure, as in c2... not just up/download data.
    And I clearly see now that this is in fact the case in III, c is not differentiated. Closing this thread for good.}

@@

  Taverna, Triana and WS-PGRADE.  \closednote{Silvia}{Should we add here the
    4?  and for translation among the 5 workflow systems in the SHIWA
    project: Askalon, MOTEUR, Taverna, Triana and WS-PGRADE)}{Tristan}{Yes, done!}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where applications, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

\subsection{Task encapsulation}
\closednote{Marc-e}{May I propose
  another name here? Either "Task-centric" or "Task Oriented". My
  reasoning is that the WE is just another task, as well as the
  sub-tasks themselves.}{Tristan}{I agree: sub-tasking is not the
  integration model, it is the process that engines use when they
  are integrated as tasks. I think 'task encapsulation' is more
  consistent  with 'service invocation' than 'task-oriented'}

@@

The science gateway also transfers both user and workflow data across
the infrastructure, so that interactions \texttt{c$_1$} and
\texttt{c$_2$} are both covered by \texttt{c}. 
\closednote{Sil}{is this next sentence relevant?}{Tristan}{I just wanted to  justify why these two interactions can be grouped but I guess you're right (I removed the sentence.}%In practice, both
%\texttt{c$_1$} and \texttt{c$_2$} can be implemented using the same
%pieces of code.

@@

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come
    here from the initial pool-model figure; it's a bit puzzling when
    mixed with letters from Fig 2}{Tristan}{Fixed.}, and workflow
  retrieval (\texttt{13} \& \texttt{14}). \closednote{Marc-e}{Maybe
    easier to follow if the following would be in fig legend instead
    of here, as for 2 previous figures}{Tristan}{fixed} Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language, which provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

\closednote{Sil}{normalization is explained in caption tab 2. i wonder whether something should be said here, as questions about the range of the score immediately pops to mind..}{Tristan}{Added a sentence about it.}

@@

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. Cell color is set as follows: (1) on each row, metric values are
  normalized between 0 (best value) and 1 (worst value):
  $m'=\frac{m-m_{\mathrm{min}}}{m_{\mathrm{max}}-m_{\mathrm{min}}}$ where
  $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
  are the minimum and maximum values for all architectures; (2) the RGB hexadecimal color code of the cell
  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
\label{table:evaluation}
\end{table*}

@@

\closednote{Sil}{i was wondering whether it would be more clear to present the metrics like this:
\item \texttt{I$_1$}: total number of components, and now the parenthesis are next to text that does not really match the metric definition (exception in this case) could also use boldface}{Tristan}{Fixed, following discussion in https://github.com/glatard/FGCS-works-special-issue/pull/20}
Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. \closednote{Sil}{add here an example of an interaction
    that is not, for example, data transfers?}{Tristan}{Done.}
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down to 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
  \closednote{Si}{new interface = new interaction? maybe better to use
    one term only here. Maybe also refer to how this shows on figure
    2}{Tristan}{Fixed the interface vs. interaction issue.}
\item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  \closednote{Silv}{it seems to me that this
    metric cannot be calculated in a straighforward manner from the
    diagrams.. is that correct? if it can, give an example of how this
    is seen in the diagram}{Tristan}{Done}
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. \closednote{sil}{again
    here it would be good to indicate how this is seen in the diagram}{Tristan}{done}
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1

>>>>

\title{Software architectures to integrate workflow engines in science gateways}

@@

\begin{abstract}
  Workflow engines are critical for the efficient and transparent
  exploitation of distributed infrastructures in the ecosystem of
  tools and services offered by science gateways.  We investigate six
  software architectures commonly used to integrate workflow engines
  into science gateways. In \emph{tight integration}, the workflow
  engine shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow import}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and support for meta-workflows and fine-grained
  debugging. Tight integration and task encapsulation are the easiest
  to integrate and the most robust. Extensibility is equivalent in
  most architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. These results provide insights for science gateway
  architects and developers.
\end{abstract}

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, or  redundancy detection across
analyses to avoid re-computation. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute
applications.

@@

Robustness is highly desirable as well since it is key to a
good user experience. Simple architectures facilitate the
implementation effort towards robust interactions which, in turn, have
a positive impact on characteristics such as gateway predictability,
transparency, reliability, traceability and
reproducibility.

Other specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, or social tools among users.

@@

@@

@@

  with heterogeneous interfaces. 

@@

  Taverna, Triana and WS-PGRADE.  

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where applications, users and other
resources are stored.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

@@

The science gateway also transfers both user and workflow data across
the infrastructure, so that interactions \texttt{c$_1$} and
\texttt{c$_2$} are both covered by \texttt{c}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3}), and workflow
  retrieval (\texttt{13} \& \texttt{14}). Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language, which provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

@@

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. Cell color is set as follows: (1) on each row, metric values are
  normalized between 0 (best value) and 1 (worst value):
  $m'=\frac{m-m_{\mathrm{min}}}{m_{\mathrm{max}}-m_{\mathrm{min}}}$ where
  $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
  are the minimum and maximum values for all architectures; (2) the RGB hexadecimal color code of the cell
  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). }
\label{table:evaluation}
\end{table*}

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down to 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.
